The design and implementation of standards-based Grid single sign-on using federated identity

### 2009-05-01 doc/sec/papers/GRID09/paper.tex
Done with proof-reading

For the service side, we implement two simple ``echoString'' services in HED and
Axis2/C. The ``echoString'' provides a simple SOAP portType which will output the same value
as it gets from input. For the client side, we develope a client by using the client API of ARC, to
invoke ``echoString'' services from Axis2/C and HED. Service invocation just sends a short
string to a service, and gets back the same string. One test involves a set of clients
which run simultaneously. Each client runs for a fixed period of time, and invokes the service as many
times as it can during the test period. Multiple clients are launched by creating multiple threads in
one test. Two values are measured: average response time and throughput. Average response
time is computed by counting the average value of the response time for each invocation.
Throughput is computed by counting number of invocations per second or per minute.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO_thread10_nosso.pdf}
\caption{Performance of HED echoString services with SAMLSSO profile: 1 service
and 10 clients}
\label{fig:SAML2SSO_thread10_nosso}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO_thread10_nosso_and_sso.pdf}
\caption{Performance of HED echoString services with SAMLSSO profile: 1 service
and 10 clients}
\label{fig:SAML2SSO_thread10_nosso_and_sso}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO_thread_all.pdf}
\caption{Performance of HED echoString services with SAMLSSO profile:
performance changes with client numbers; session non-reused }
\label{fig:SAML2SSO_thread_all}
\end{figure}

Figure \ref{fig:SAML2SSO_thread10_nosso} depicts the average response time and
throughputs for 10 clients running simultaneously against the ``echoString'' service enhanced with SAML2SSO
profile. Since the success of authentication from IdP can be regarded as valid for the whole
session between client and service, the client with ``Session reused'' need to accomplish the SAML2SSO
profile only once for all SOAP invocations on this client, while the client with ``Session
non-reused'' will need to accomplish this profile once per SOAP invocation.

The average response time of ``Session non-reused'' experiments changes little
with the execution duration, and the value is about 3000ms; while the throughput is around 200
requests per minutes. Considering that the average response time in case of ``Session non-reused'' and
pure SOAP invocation is around 15ms when client number is 10 (see section \ref{sec:perhedandaxis}),
therefore we can calculate that the consumption of time for SAML2SSO profile is about 99.5\% of
the whole SOAP invocation with SAML2SSO profile integrated.

Deeper investigation about recording the time of each step of SAML2SSO profile
supports the above observation. The average time cost for getting an authentication request
from SP service is about 60ms; the average time cost for authenticating against IdP is about
2750ms; the average time cost for sending the SAML response (from IdP) to SP service is
about 20ms; the average time cost for verifying (by SP service) the signature of SAML
response is around 130ms.

In order to improve the performance of SAML2SSO profile, optimizing the process
between client and IdP is the correct direction, and probably one of the easiest way is
to enhance the hardware configuration.

One the other hand, the average response time of ``Session reused'' experiments
is about 82 ms, which is almost as short as the value in the pure SOAP invocation experiment in
section \ref{sec:perhedandaxis}. The result is reasonable because the time consumption
caused by only once of SAML2SSO execution on one client is relatively small comparing with
the time consumption caused by the following hundreds of pure SOAP invocations on the
same client.

Figure~\ref{fig:SAML2SSO_thread_all} depicts the performance results for
echoString services that are enhanced with SAML2SSO profile, with client number being changed. In
this set of experiments, the session is not reused, and secure communication without client certificate
authentication is used. The results show the average response time increases linearly with the client
number, while the throughput increases initially and then keeps in a fixed value (around 215
requests per minute) when the client number is bigger than 4. When the client number is four, the
response time is around one second, while the throughput of the service is 215 requests per
minutes which is almost the fixed value.
We speculate that in order to get acceptable response time, there should be less
than four (including four) concurrent clients if they need to continuously participate one SAML2SSO
profile.

Considering that the throughput of the ``echoString'' service enhanced with mutual
TLS is 13000, and the concurrent client number when arriving fixed throughput is 30, the
performance of ``echoString'' service enhanced with SAML2SSO profile is much worse, with the
values being 215 and 4 correspondingly.

However for some typical grid application scenarios, normally the client does
not need to continuously accomplish SAML2SSO profile. For instance, a user could submit a
job which needs a period of time (normally this period is much longer than a few
seconds) to execute, which launches an execution of SAML2SSO profile; 
afterwards, this user could query the information as well as the result about
this job, which launches another execution of SAML2SSO profile. On the other hand,
statistically, users will most probably not access grid system in a concurrent way, rather than a random way.
Therefore, if we suppose a user could access grid once per minute, in our test environment, we
can claim that around 215 clients randomly can be supported with acceptable performance.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SLCS_thread_all.pdf}
\caption{Performance of SLCS service: performance changes with client numbers}
\label{fig:SLCS_thread_all}
\end{figure}

Figure \ref{fig:SLCS_thread_all} is the performance about the SLCS service.
Since the SLCS is based on the SAML2SSO profile, the response time should be the accumulation of
time consumption about the SAML2SSO processing and the SLCS processing. In this experiment, SLCS
service and client run on the same set of machine as section~\ref{sec:perhedandaxis}.

The results show that the throughput of SLCS service has the same characteristic
as the throughput of ``echoString'' service enhanced with SAML2SSO profile, with the value being
around 215 requests per minute; and the average response time resembles the value of
``echoString'' service as well, with the value being increased for about 80ms. The increased time
consumption is caused by the creating of X.509 request (1024-bit RSA key pair generation) on the client
side.

@@

In the last experiment, we evaluate how does the usage of delegation service
impact on the performance of Web Service invocation. We deployed a ``echoString'' service
plugged with delegation security handler on a machine equipped with 3GHz Intel Pentium dual-core CPU and
2GB memory, Red Hat Enterprise Linux 5, and also configure a client together with delegation
security handler plugged on a machine equipped with 2.0GHz AMD Athlon CPU and 1GB memory, Red Hat
Enterprise Linux 5. Aside from ``echoString'' service and client, a delegation service is deployed on a machine
equipped with 3GHz Intel Pentium dual-core CPU and 2GB memory, Red Hat Enterprise Linux 5.

\begin{figure}
\includegraphics[width=0.9\columnwidth]{Delegation_thread_to_perf.pdf}
\caption{Performance of HED echoString service with credential delegation:
performance changes with client numbers; echoString service contacts delegation service upon
each service invocation from client}
\label{fig:Delegation_thread_to_perf}
\end{figure}

In the experiment, the client launches a new credential delegation for each
service invocation, and ``echoString'' service acquires delegated credential from delegation service
upon each service invocation.
Figure~\ref{fig:Delegation_thread_to_perf} depicts the average response time and
throughput of the ``echoString'' service with delegation security handler embedded. The
average response time increases linearly with the client number, while the throughput starts to keep at
fixed value (around 400 request per minutes) when the client number is 8, after initially
increasing.
We conclude that if the client number is bigger than eight, in our test
environment, there is no throughput benefit can be achieved by adding clients.

%-------------------------------------------------------------------------
\section{Related Work}
\label{sec:relatedwork}
There are other researches that utilize Shibboleth for protecting usage of Grid
resources. The work in ~\cite{J.Watt06} uses Shibboleth to protect a GridSphere portal and
then indirectly protect Grid services which are deployed in this GridSphere portal.
Another work in ~\cite{R.O.Sinnott06} also uses Shibboleth to protect Grid portal,
moreover, it maintains a few common used server certificates which can then be used to access
Grid resources after users have successfully authenticated themselves at Grid portal
through Shibboleth. Unlike these two solutions that use Shibboleth to protect Grid
portal, the solution in this paper uses Shibboleth or similar standards-based identity federation to
directly protect Grid resource/services, which provide possibility for those Grid applications
that are directly based on Grid interface (e.g., applications that invokes SOAP operations of
remote Grid services), to benefit from identity federation.

GridShib~\cite{T.Scavo07,VWelch05} provides one solution that users can use
X.509 certificate to authenticate to Shibboleth IdP, get back SAML assertion and embed
it in the proxy certificate. The Grid service will then use this SAML assertion for
access control. Unlike the work described in this paper, GridShib still requires users to
possess a X.509 credential, rather than username/password. GridShib also provides another 
solution~\cite{TBarton06} in which a credential service (MyProxy server) is
deployed together with Shibboleth IdP for acting as an online-CA, authenticating the user through
username/password rather than X.509 certificate, and issuing short-lived X.509 credential. The
second solution is similar to the short-lived credential service in our solution, while our
solution also provides the option for users directly authenticating to Grid services via
SAML2.0 SSO profile using their username/password.

Another short-lived credential service is also provided by SWITCH~\cite{switchslcslink}. 
The advantage of our work comparing to the work of SWITCH is that 
our SLCS service is based on the Web Services standard, thus that it is easier to
achieve interoperability.

@@

%-------------------------------------------------------------------------
\section{Conclusion And Future Work}
\label{sec:conclusion}
In this paper we propose a Grid single sign-on framework that can utilize
users' federated identity instead of X.509 mutual authentication for authentication,
which we believe can make further uptake of Grid resources as well as Grid middleware
by wider scale of users.

We implement the SAML2 SSO profile in the ARC Grid middleware, so that the Grid
applications that developed by using the application interface provided by this middleware
can easily be switched to benefit from federated identity based authentication. 
The performance evaluation shows even though the introducing of SAML2 SSO
profile causes much performance downgrade, this downgrade can be avoided by reusing the
authentication result for multiple SOAP invocations.
Meanwhile, in order to be compatible with those Grid applications that require
X.509 certificate on the client side, we also implement one Web Service which can be
used for obtaining an X.509 credential based on federated identity, and the other Web
Service which can be used for delegating X.509 credential. We believe that the
evaluated performance is enough for creating X.509 certificate and proxy certificate,
considering that the certificates should have a relatively much longer lifetime comparing to the
response time of these two services.

Although only authentication issue has been discussed in this paper, we know
single sign-in should also include the authorization issue. The SAML assertion which
includes users' attributes can be used on the service side to achieve attribute-based
authorization. Therefore providing plug-in on the service side to enforce access control using
SAML attribute is considered one part of the future work.

SimpleSAMLphp~\cite{simplesamllink} provides another implementation of SAML2 SSO
profile, and it has been widely deployed around Nordic countries for identity federation. Hence
integrating ARC based Grid applications with simpleSAMLphp is considered as another of
future work.

>>>>

For the service side, we implement two simple ``echoString'' services in HED and
Axis2/C. The ``echoString'' provides a simple SOAP portType which will output the same value
as it gets from input. For the client side, we develop a client by using the client API of ARC, to
invoke ``echoString'' services from Axis2/C and HED. Service invocation just sends a short
string to a service, and gets back the same string. One test involves a set of clients
which run simultaneously. Each client runs for a fixed period of time, and invokes the service as many
times as it can during the test period. Multiple clients are launched by creating multiple threads in
one test. Two values are measured: average response time and throughput. Average response
time is computed by counting the average value of the response time for each invocation.
Throughput is computed by counting number of invocations per second or per minute.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO_thread10_nosso.pdf}
\caption{Performance of HED echoString services with SAML2SSO profile: 1 service
and 10 clients; single sign-on is disabled; session used and session not re-used are compared}
\label{fig:SAML2SSO_thread10_nosso}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO_thread10_nosso_and_sso.pdf}
\caption{Performance of HED echoString services with SAML2SSO profile: 1 service
and 10 clients; session not re-used; single sign-on enabled and single sign-on disabled are compared}
\label{fig:SAML2SSO_thread10_nosso_and_sso}
\end{figure}

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO_thread_all.pdf}
\caption{Performance of HED echoString services with SAML2SSO profile:
performance changes with client numbers; session not re-used; single sign-on enabled and 
single sign-on disabled are compared}
\label{fig:SAML2SSO_thread_all}
\end{figure}

We consider two scenarios of authenticating between user agent (client) and IdP: 
SSO enabled, and SSO disabled. ``SSO disabled'' means the successful authentication
session between user agent and IdP is not regarded as valid for the subsequent authentication, and
user is required to authenticate by inputing his username and password on each request.
``SSO enabled'' means the successful authentication session is regarded as valid for the 
subsequent authentication during a short period (we set up this period at 30 minutes).

Since the success of authentication from IdP can be regarded as valid for the whole
session between client and service, we also consider two scenaros in 
usage of the authentication result between client and service: session re-used, and session non-reused.
The client with ``Session reused'' need to accomplish the SAML2SSO
profile only once for all SOAP invocations on this client, while the client with ``Session
non-reused'' will need to accomplish this profile once per SOAP invocation.

Figure~\ref{fig:SAML2SSO_thread10_nosso} shows the average response time and
throughputs for 10 clients running simultaneously against the ``echoString'' service enhanced with SAML2SSO
profile, with single sign-on disabled.

The average response time of ``Session non-reused'' experiments changes little
with the execution duration, and the value is about 3000ms; while the throughput is around 215
requests per minute. Given that the average response time in case of ``Session non-reused'' and
pure SOAP invocation is around 15ms for 10 clients (see Section~\ref{sec:perhedandaxis}),
we can evaluate that the consumption of time for SAML2SSO profile is about 99.5\% of
the whole SOAP invocation with SAML2SSO profile integrated.

Deeper investigation involving recording the time for each step of SAML2SSO profile
supports the above observation. The average time cost for getting an authentication request
from SP service is about 60ms; the average time cost for authenticating against IdP is about
2750ms; the average time cost for sending the SAML response (from IdP) to SP service is
about 20ms; the average time cost for verifying (by SP service) the signature of SAML
response is around 130ms.

In order to improve the performance of SAML2SSO profile, optimizing the process
between client and IdP is the correct direction, and probably one of the easiest ways is
to enhance the hardware configuration.

One the other hand, the average response time of ``Session reused'' experiments
is about 82 ms, which is almost as short as the value in the pure SOAP invocation experiment in
Section~\ref{sec:perhedandaxis}. The result is reasonable because the time consumption
caused by only once of SAML2SSO execution on one client is relatively small comparing with
the time consumption caused by the following hundreds of pure SOAP invocations on the
same client. So besides improving the performance of SAML2SSO profile, reusing 
the session as long as possbile is also a way to improve the performance of 
Web Service invocation with SAML2SSO profile integrated as authentication approach.

Figure~\ref{fig:SAML2SSO_thread10_nosso_and_sso} depicts performance comparison between single 
sign-on disabled and single sign-on enabled scenarios, for 10 clients running simultaneously 
against the ``echoString'' service enhanced with the SAML2SSO profile, with session not re-used.

The average response time in the ``SSO enabled'' case is about 1700ms, while the throughput is about
340 requests per minutes. The average response time in the ``SSO disabled'' case is about 3000ms, while 
the throughput is about 215 requests per minutes. The comparison shows that the performance 
of ``SSO enabled'' is better than that of ``SSO disabled'', but still on the same quantitative
level.

Figure~\ref{fig:SAML2SSO_thread_all} illustrates performance results for
echoString services that are enhanced with SAML2SSO profile, with client number being changed. In
this set of experiments, the session is not re-used, and single sign-on enabled and single sign-on
disabled scenarios are compared.

For the situation when single sign-on is disabled, the average response time 
increases linearly with the number of clients, while the throughput increases initially and then 
stays constant (around 215 requests per minute) when the client number is bigger 
than 4. When the client number is 4, the response time is around one second, while the 
throughput of the service is 215 requests per minute, which is almost constant.

For the situation of single sign-on enabled, the average response time 
also increases linearly with the client number, while the throughput also increases initially and then 
saturates at around 340 requests per minute when the client number is bigger 
than 5. When the client number is five, the response time is around 880ms, while the 
throughput of the service is 340 requests per minute, which is almost constant.

We speculate that in order to get acceptable response time, for the case when single sign-on is disabled, 
there should be less than four (including four) concurrent clients if they need to continuously 
participate one SAML2SSO profile; for the case when single sign-on is enabled, 
there should be less than five (including five) concurrent clients if they need to continuously 
participate one SAML2SSO profile.

Given that the throughput of the ``echoString'' service enhanced with mutual
TLS is 13000, and the concurrent client number when throughput reaches saturation is 30, the
performance of ``echoString'' service enhanced with SAML2SSO profile is much worse, both 
with single sign-on disabled and enabled, with the values being 215, 4, and 340, 5, correspondingly.

However, for some typical grid application scenarios, the client normally does
not need to continuously accomplish SAML2SSO profile. For instance, a user could submit a
job which needs a period of time (normally this period is much longer than a few
seconds) to execute, which launches an execution of SAML2SSO profile. 
This user could afterwards query the information as well as the result of
this job, which would launch another execution of SAML2SSO profile. On the other hand,
statistically, users will most probably not access grid system in a concurrent way, rather than in a random way.
Therefore, if we suppose that a user could access grid once per minute, in our test environment, we
can claim that for the case of disabled single sign-on, around 215 clients randomly can be 
supported with acceptable performance, while for the case of single sign-on enabled, around 340 
clients randomly can be supported with acceptable performance.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SLCS_thread_all.pdf}
\caption{Performance of SLCS service: performance changes with client numbers; single sign-on disabled}
\label{fig:SLCS_thread_all}
\end{figure}

Figure~\ref{fig:SLCS_thread_all} shows performance of the SLCS service.
Since the SLCS is based on the SAML2SSO profile, the response time should be the combination of
time consumption for the SAML2SSO processing and the SLCS processing. In this experiment, SLCS
service and client run on the same set of machine as in Section~\ref{sec:perhedandaxis}.

Since the time interval of the usage of an SLCS client should be similar to the lifetime of the SLCS
credential, which is 12 hours by default, and the lifetime of a successful authentication session is 
30 minutes by default, enabling single sign-on is not useful in a real deployment because a successful 
authentication session will expire before the next time the SLCS client is used. Therefore, 
we do not enable single sign-on when testing the SLCS service.

The results show that the throughput of SLCS service has the same characteristic
as the throughput of ``echoString'' service enhanced with SAML2SSO profile, with the value being
around 215 requests per minute; and the average response time is close to the value of
``echoString'' service as well, with the value being increased to about 80ms. The increased time
consumption is caused by the creating of X.509 request (1024-bit RSA key pair generation) on the client
side.

@@

In the last experiment, we evaluate how does the usage of delegation service
affect performance of Web Service invocation. We deployed an ``echoString'' service
plugged with delegation security handler on a machine equipped with 3GHz Intel Pentium dual-core CPU and
2GB memory, Red Hat Enterprise Linux 5, and also configured a client together with delegation
security handler plugged on a machine equipped with 2.0GHz AMD Athlon CPU and 1GB memory, Red Hat
Enterprise Linux 5. Aside from ``echoString'' service and client, a delegation service was deployed on a machine
equipped with 3GHz Intel Pentium dual-core CPU and 2GB memory, Red Hat Enterprise Linux 5.

\begin{figure}
\includegraphics[width=0.9\columnwidth]{Delegation_thread_to_perf.pdf}
\caption{Performance of HED echoString service with credential delegation:
performance changes with the number of clients; echoString service contacts delegation service upon
each service invocation from client}
\label{fig:Delegation_thread_to_perf}
\end{figure}

In this experiment, the client launches a new credential delegation for each
service invocation, and ``echoString'' service acquires delegated credential from delegation service
upon each service invocation.
Figure~\ref{fig:Delegation_thread_to_perf} depicts the average response time and
throughput of the ``echoString'' service with delegation security handler embedded. The
average response time increases linearly with the client number, while the throughput starts to saturate
at around 400 request per minutes when the client number reaches 8.
We conclude that if the client number is bigger than eight, in our test
environment, there is no throughput benefit can be achieved by adding clients.

%-------------------------------------------------------------------------
\section{Related Work}
\label{sec:relatedwork}

There are other efforts that utilize Shibboleth for protecting usage of Grid
resources. The work of Watt and colleagues~\cite{J.Watt06} uses Shibboleth to protect a GridSphere portal and
then indirectly protect Grid services which are deployed in this GridSphere portal.
Another work of Sinott and colleagues~\cite{R.O.Sinnott06} also uses Shibboleth to protect a Grid portal,
moreover, it maintains a few commonly used server certificates which can then be used to access
Grid resources after users have successfully authenticated themselves at a Grid portal
through Shibboleth. Unlike these two solutions that use Shibboleth to protect Grid
portal, the solution in this paper uses Shibboleth or similar standards-based identity federation to
directly protect Grid resource/services, which provides possibility for those Grid applications
that are directly based on Grid interface (e.g., applications that invoke SOAP operations of
remote Grid services), to benefit from identity federation.

GridShib~\cite{T.Scavo07,VWelch05} provides a solution where users can use
X.509 certificates to authenticate to Shibboleth IdP, get back SAML assertion and embed
it in the proxy certificates. The Grid service will then use this SAML assertion for
access control. Unlike the work described in this paper, GridShib still requires users to
possess a X.509 credential, rather than username/password. GridShib also provides another 
solution~\cite{TBarton06} in which a credential service (MyProxy server) is
deployed together with Shibboleth IdP for acting as an online-CA, authenticating the user through
username/password rather than X.509 certificate, and issuing short-lived X.509 credential. The
second solution is similar to the short-lived credential service in our work, while our
solution also provides the option for users to directly authenticate to Grid services via
SAML2.0 SSO profile using their username/password.

Another short-lived credential service is also provided by SWITCH~\cite{switchslcslink}. 
The advantage of our work comparing to the work of SWITCH is that 
our SLCS service is based on the Web Services standard, thus it is easier to
achieve interoperability.

@@

%-------------------------------------------------------------------------
\section{Conclusion And Future Work}
\label{sec:conclusion}
In this paper we propose a Grid single sign-on framework that can utilize
users' federated identity instead of X.509 mutual authentication for authentication,
which we believe can enhance uptake of Grid technlogies by resources owners as well as
by wider range of users.

We implement the SAML2 SSO profile in the ARC Grid middleware, so that the Grid
applications developed by using the application interface provided by this middleware
can easily be adapted to benefit from the federated identity based authentication. 
The performance evaluation shows that even though introduction of SAML2 SSO
profile causes notable performance downgrade, this downgrade can be avoided by reusing the
authentication result for multiple SOAP invocations.
Meanwhile, in order to be compatible with those Grid applications that require
X.509 certificate on the client side, we also implement one Web Service which can be
used for obtaining an X.509 credential based on federated identity, and another Web
Service which can be used for delegating X.509 credential. We believe that the
observed performance is sufficient for creating X.509 certificates and proxy certificates,
considering that the certificates should have a relatively much longer lifetime comparing to the
response time of these two services.

Although only the authentication issue has been discussed in this paper, we know that
single sign-in should also include the authorization aspect. The SAML assertion which
includes users' attributes can be used on the service side to achieve attribute-based
authorization. Therefore providing plug-in on the service side to enforce access control using
SAML attribute is considered an important part of the future work.

SimpleSAMLphp~\cite{simplesamllink} provides another implementation of SAML2 SSO
profile, and it has been widely deployed e.g. around Nordic countries for identity federation. Hence
integrating ARC based Grid applications with simpleSAMLphp is considered as another of
future work.
